---
title: "R Notebook"
output: html_notebook
---

```{r}
#Append conda environment bin path to PATH
conda.env.bin.path.assembly_annotation <- "/home/troyalty/Software/bioinformatics/miniconda3/envs/assembly_annotation/bin"

#There seems to be a conflict between bowtie2 and some other installations in assembly_annotation.
#I created a new conda environment that only contains bowtie2. This isn't the prettiest solution,
#but seems to work.
conda.env.bin.path.bowtie2_only <- "/home/troyalty/Software/bioinformatics/miniconda3/envs/bowtie2_only/bin"
Sys.setenv(PATH = paste(Sys.getenv("PATH"), conda.env.bin.path.assembly_annotation,conda.env.bin.path.bowtie2_only, sep = ":")) 

```

```{r}
knitr::opts_chunk$set(engine.opts = list(bash = "-l"))

library(tidyverse)
library(foreach)
library(doParallel)

source("scripts/R/assembly_annotation_functions.R")
source("scripts/R/postprocessing_functions.R")
```

## Preprocessing pipeline 

This pipeline is designed to isolate putative genes from shotgun metagenomic sequence datasets, annotate target genes using dbCAN, and isolate putatively exported proteins. Most functions implement shell commands via subprocesses and require all softwares to be included in the system PATH. In my case, I created a conda environment which includes all the softwares below (with the exception of signalp), and appended that conda environment's bin to my PATH.

The workflow is:
1) append conda environment bin path to PATH
2) create standardized directory structure
3) download short-read datasets from SRA using fastq-dump
4) quality filter short-reads using fastp
5) assemble high-quality reads into contigs using megahit
6) predict putative genes using prodigal
7) create a hmm database using dbCAN definitions
8) annotate genes with dbCAN definitions using HMMER
9) isolate dbCAN-positive genes
10) separate putative extracellular and intracellular enzymes based on signalp predictions--cleave signal peptides

```{r warning=FALSE}
accession.list <- c("SRR7066493","SRR7066492")
override <- TRUE
fastq.dump.cores=2
megahit.cores=30
min.contig.length <- 1000
dbcan.db <- 'data/database/dbCAN-HMMdb-V10.txt'


#create standardized directory
create_subdirectories(override = override)

#download specified short read datasets
download_SRR(accession.list = accession.list, 
             numCores = fastq.dump.cores)

#trim adaptars and drop low quality reads
trim_reads(accession.list = accession.list)

#assemble short reads into contigs
assemble_reads(accession.list, 
               megahit.cores=megahit.cores, 
               min.contig.length = min.contig.length)

#predict contig ORFs
predict_ORFs(accession.list)

#make a dbcan database
make_database(dbcan.db)

#annotate genes with dbcan annotations
dbcan_annotation(accession.list = accession.list, 
                 db=dbcan.db)

#filter and isolate genes annotated with dbcan annotations  
filter_fasta_with_gene_list(accession.list = accession.list, 
                            dir.gene.list = "data/sequence/annotation/hmm/", 
                            ex_gene_list = "_dbCAN-HMMdb-V10.txt_parsed.tsv", 
                            col = 3)

#separate dbCAN-positive annotations in signal peptide-positive and negative sequences--cleave signal peptide
filter_signal_peptide(accession.list = accession.list, 
                      faa_ex="_filtered.faa", cleave = TRUE)

```

### Isolate dbCAN Enzymes with Largest Richness

```{r warning=FALSE}
#sub-sample

dir.hmm.db <- "data/sequence/annotation/hmm/"
hmm.gene.list <- "_dbCAN-HMMdb-V10.txt_parsed.tsv"
dir.extracellular<-'data/sequence/genes/aa/signalp/extracellular/'
dir.blast.output <- 'data/sequence/annotation/blastp/'
db.path <- 'data/database/'
target_ex <- "_extracellular_filtered.faa"

blast.options <- "-max_target_seqs 1 -evalue 0.1 -outfmt 6"
n_sample <- 40
sample.accession <- "SRR7066493"
blast.accession.list <- c("SRR7066492")
hmm.list <- c("GH109.hmm","GH29.hmm","GT2_Glycos_transf_2.hmm")
blast.cores <- 30



#create directory for genes that we will predict protein structures for the most rich proteins 
dir.struct.rich <- 'data/sequence/structures/genes/richness_analysis/'
if (dir.exists(dir.struct.rich)){
  unlink(dir.struct.rich,recursive = TRUE)
}
dir.create(dir.struct.rich,recursive = TRUE)

#Identify enzyme annotations with greatest richness
plot_rank_abundance(accession.list,
                    dir=dir.hmm.db,
                    ex=hmm.gene.list,
                    header=FALSE,
                    name.split="_")


#Isolate genes with highest richness
for (hmm in hmm.list){
  hmm.short<-strsplit(hmm,'[.]')[[1]][1]
  hmm.ex.tmp=paste0("_",hmm.short,'.faa')
  
  
  filter_fasta_with_gene_list(accession.list,
                              dir.gene.list=dir.hmm.db,
                              dir.genes=dir.extracellular,
                              ex_gene_list=hmm.gene.list,
                              ex_gene=target_ex,
                              out_ex=hmm.ex.tmp,
                              col=3,
                              apply.grep = TRUE,
                              grep.filter = hmm)
  
  sample_fasta(accession.list = sample.accession, 
               n = n_sample,
               gene_ex = hmm.ex.tmp,
               dir.gene = dir.extracellular)
  
  
  for (accession in blast.accession.list){
    #copy sampled file to data/database to create blast database
    system(sprintf("cp data/sequence/genes/aa/signalp/extracellular/%s_extracellular_filtered.faa %s",
                   accession,
                   db.path))
    
    #make blast database
    db.path.tmp=paste0(db.path,accession,"_extracellular_filtered.faa")
    make_database(file=db.path.tmp,fun = "makeblastdb")
    
    
    #blast other files to find homolog pairings
    blastp_query(sample.accession, 
                 db.path=db.path.tmp,
                 dir.output=dir.blast.output,
                 dir.genes=dir.extracellular,
                 genes_ex=paste0("_",n_sample,hmm.ex.tmp),
                 blast.options=blast.options, 
                 blast.cores=blast.cores)
    
    blast_ex<-paste0("_",n_sample,"_",hmm.short,"_hits.blastp")
    
    system(sprintf("mv data/sequence/annotation/blastp/%s.blastp data/sequence/annotation/blastp/%s%s",
                   sample.accession,
                   accession,
                   blast_ex))
    
    #filter homologous proteins and place in structure prediction directory
    filter_fasta_with_gene_list(accession,
                                dir.gene.list=dir.blast.output,
                                dir.genes=dir.extracellular,
                                ex_gene_list=blast_ex,
                                ex_gene=target_ex,
                                out_ex=paste0("_",n_sample,"_",hmm.short,"_hits.faa"),
                                col=2) 
    
  }
  
  #move target genes to directory for structure prediction
  system(sprintf("mv data/sequence/genes/aa/signalp/extracellular/*%s_%s* %s",
                 n_sample,
                 hmm.short,
                 dir.struct.rich))
  
}


```

### Isolate the most abundant dbCAN Enzymes based on short-read mapping 

```{r warning=FALSE}

dir.nuc.extracellular <- 'data/sequence/genes/nuc/signalp/extracellular/'
dir.aa.extracellular <- "data/sequence/genes/aa/signalp/extracellular/"
dir.genes.nuc.tmp <- 'data/sequence/genes/nuc/signalp/extracellular/tmp/'
dir.genes.nuc <- 'data/sequence/genes/nuc/'
ex_gene_list <- '_abundant.txt'
db.path <- 'data/database/'
dir.blast.output <- 'data/sequence/annotation/blastp/'
dir.hmm.db <- "data/sequence/annotation/hmm/"

hmm.gene.list <- "_dbCAN-HMMdb-V10.txt_parsed.tsv"
out.ex <- "_extracellular_filtered.fna"
in.aa.ex <- "_extracellular_filtered.faa"
out.aa.ex <- "_extracellular_abundant.faa"
in.nuc.ex <- ".fna"
accession.list <- c("SRR7066493","SRR7066492")


blast.options <- "-max_target_seqs 1 -evalue 0.1 -outfmt 6"
n_sample <- 40
sample.accession.list <- "SRR7066492"
blast.accession <- "SRR7066493"
blast.cores <- 30

#create directory for genes that we will predict protein structures for the most abundant proteins 
dir.struct.RA <- 'data/sequence/structures/genes/RA_analysis/'

if (dir.exists(dir.struct.RA)){
  unlink(dir.struct.RA,recursive = TRUE)
}
dir.create(dir.struct.RA,recursive = TRUE)


if(dir.exists(dir.nuc.extracellular)){
  unlink(dir.nuc.extracellular, recursive = TRUE)
}
dir.create(dir.nuc.extracellular, recursive = TRUE)

if(dir.exists(dir.genes.nuc.tmp)){
  unlink(dir.genes.nuc.tmp, recursive = TRUE)
}
dir.create(dir.genes.nuc.tmp,recursive = TRUE)


# aa.tmp <- paste0(dir.aa.extracellular,blast.accession,in.aa.ex)
aa.tmp <- paste0(dir.aa.extracellular,blast.accession,in.aa.ex)
nuc.tmp <- paste0(dir.genes.nuc,blast.accession,in.nuc.ex)
list.tmp <- paste0(dir.genes.nuc.tmp,blast.accession,'_list.txt')
out.nuc.tmp <- paste0(dir.nuc.extracellular,blast.accession,out.ex)


# cmd.create.list <- sprintf("grep \"^>\" %s | cut -f 1 -d ' ' | sed 's/>//' > %s",
#                            aa.tmp,
#                            list.tmp)

cmd.create.list <- sprintf("cut -f 3 %s > %s",
                           paste0(dir.hmm.db,blast.accession,hmm.gene.list),
                           list.tmp)

system(cmd.create.list)


cmd.extract <- sprintf("seqtk subseq %s %s > %s",
                       nuc.tmp,
                       list.tmp,
                       out.tmp)


system(cmd.extract)


#align high quality short reads to tentatively exported dbcan positive enzymes
align_short_reads(blast.accession, 
                  dir.genes.nuc=dir.nuc.extracellular,
                  ex_gene=out.ex,
                  align.cpu=30)

#predict ORFs for signal peptide positive and dbcan annotated genes--this is to generate a gff for conversion to gtf in alignment_to_abundance  
predict_ORFs(blast.accession,gff.only=TRUE,dir.assembly = dir.nuc.extracellular, in.ex=out.ex, out.ex=out.ex, extension=out.ex)

alignment_to_abundance(blast.accession,ex.gff=paste0(out.ex,".gff"), ex.gtf=paste0(out.ex,".gtf"), gff2gtf=TRUE) 




#plot rank abundance curve
df <- rank_relative_abundance(blast.accession,
                              dir='data/sequence/genes/abundance/',
                              ex="_abundance.tsv",
                              header=FALSE,
                              name.split="_",
                              skip=2,
                              col = c(2,6,7))

#write a temporary list to isolate amino acid sequences
abundant.list <- paste0(dir.genes.nuc.tmp,blast.accession,ex_gene_list)
data.table::fwrite(as.list(df$V2),paste0(dir.genes.nuc.tmp,blast.accession,ex_gene_list),col.names = FALSE, row.names = FALSE,quote = FALSE,sep='\n')

#filter most abundant genes from faa file
filter_fasta_with_gene_list(blast.accession,
                            dir.gene.list=dir.genes.nuc.tmp,
                            dir.genes=dir.aa.extracellular,
                            ex_gene_list=ex_gene_list,
                            ex_gene=in.aa.ex,
                            out_ex=out.aa.ex,
                            col=1)


#copy most abundant file for blast analysis to data/database to create blast database
for (a in sample.accession.list){
  system(sprintf("cp data/sequence/genes/aa/signalp/extracellular/%s%s %s",
                 a,
                 in.aa.ex,
                 db.path))
  
  #make blast database
  db.path.tmp=paste0(db.path,a,in.aa.ex)
  make_database(file=db.path.tmp,fun = "makeblastdb")
  
  #blast other files to find homolog pairings
  blastp_query(blast.accession, 
               db.path=db.path.tmp,
               dir.output=dir.blast.output,
               dir.genes=dir.aa.extracellular,
               genes_ex=out.aa.ex,
               blast.options=blast.options, 
               blast.cores=blast.cores)
  
  blast_ex<-"_abundant_hits.blastp"
  
  system(sprintf("mv data/sequence/annotation/blastp/%s.blastp data/sequence/annotation/blastp/%s%s",
                 blast.accession,
                 a,
                 blast_ex))
  
  
  #filter homologous proteins and place in structure prediction directory
  filter_fasta_with_gene_list(a,
                              dir.gene.list=dir.blast.output,
                              dir.genes=dir.aa.extracellular,
                              ex_gene_list=blast_ex,
                              ex_gene=target_ex,
                              out_ex=out.aa.ex,
                              col=2) 
  
  #move target genes to directory for structure prediction1
  system(sprintf("cp %s%s%s %s",
                 dir.aa.extracellular,
                 a,
                 out.aa.ex,
                 dir.struct.RA))
  
  
  
}

#move target genes to directory for structure prediction1
system(sprintf("mv data/sequence/genes/aa/signalp/extracellular/%s%s %s",
               blast.accession,
               out.aa.ex,
               dir.struct.RA))


```

# Run alphafold2 on isolated genes


```{bash echo=FALSE}
#filepath containing fasta files
faa_filepath=/home/troyalty/Documents/projects/alphafold_stability/data/sequence/structures/genes/richness_analysis
#filepath housing alphafold2 run_docker.py script
alphafold2_script=/home/troyalty/Software/alphafold2_docker/alphafold/docker/run_docker.py
#directory to output structure results
output_path=/home/troyalty/Documents/projects/alphafold_stability/data/sequence/structures/genes/richness_analysis/structures
#a temporary directory for writing temporary fasta files
tmp_path=/home/troyalty/Documents/projects/alphafold_stability/data/sequence/structures/tmp/
#directory containing alphafold2 reference data
reference_data=/srv/data/alphafold/

#this allows for using conda's activate command. Note, that conda's path is added to the PATH in ~/.profile. There is apparently some variability in how
#rstudio executes .bashrc, .bash_profile, .profile, etc. among different OS. GL...
source ~/.profile
source activate
conda activate alphafold_docker

rm -fr $tmp_path 2> /dev/null
mkdir $tmp_path

rm -fr $output_path 2> /dev/null
mkdir $output_path


#loop through list 
for f in $faa_filepath/*.faa; do
cat $f | while read -r header; do #the while loop iterates through headers and sequences--note this requires the fasta file to be single-line fasta format
read -r seq #this reads the sequence under the header
gene_name=$(echo $header | cut -f 1 -d ' ' | cut -f 2 -d '>')
seq_file=$tmp_path/$gene_name$fa_ex

#alphafold only takes one sequence at at time; this requires isolating sequences in multifasta files into temporary files
echo $header > $seq_file
echo $seq >> $seq_file

#run the alphafold2 run_docker.py script
python3 $alphafold2_script \
--fasta_paths=$seq_file \
--max_template_date=2020-05-14 \
--data_dir=$reference_data \
--output_dir=$output_path
done
done

rm -fr $tmp_path 2> /dev/null
```


